---
title: "Alternative_promoter"
author: "Quynh Nhu Nguyen"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3   # Optional: Specify the depth of headers to include in the TOC
  pdf_document:
    toc: true
    toc_depth: 3   # Optional: Specify the depth of headers to include in the TOC
---

# A complete workflow to identify alternative promoter usage

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Load the necessary libraries

```{r}
library(proActiv)   # For promoter analysis
library(gridExtra)  # For arranging multiple plots
library(Rtsne)      # For t-SNE dimensionality reduction
library(ggplot2)    # For creating plots and visualizations
library(SummarizedExperiment)
library(tidyr)
library(biomaRt)

library(dplyr)
```

## Preparing promoter annotations

```{r}
## Define the GTF file path for promoter annotation
gtf_file_hg19 <- '/mnt/NAS4/user_data/np-nhu/rna_human/genome/gencode.v19.annotation.gtf.gz'
file.exists(gtf_file_hg19) # Check if the GTF file exists

## Prepare promoter annotation using the GTF file for Homo sapiens
promoterAnnotation_gencode_v19 <- preparePromoterAnnotation(
    file = gtf_file_hg19, 
    species = "Homo_sapiens")

## Define the TxDb file path for promoter annotation
txdb_file_hg19 <- '/mnt/NAS4/user_data/np-nhu/rna_human/genome/gencode.v19.annotation.sqlite'
file.exists(txdb_file_hg19) # Check if the TxDb file exists

## Load the TxDb object from the SQLite file
txdb_hg19 <- loadDb(txdb_file_hg19)

## Prepare promoter annotation using the TxDb object for Homo sapiens
promoterAnnotation_gencode_v19 <- preparePromoterAnnotation(
    txdb = txdb_hg19, 
    species = "Homo_sapiens")
```

## Preparing input data

```{r}
## Define the directory path containing STAR junction files
directory_path <- '/mnt/NAS4/user_data/np-nhu/rna_human/junctions_filter_17092024_c25'

## List all files in the directory with full paths
files <- list.files(directory_path, full.names = TRUE)
files # Display the listed files

## Define the experimental conditions for the samples
condition <- rep(c("HIGH", "LOW"), times = c(8, 13))
condition
```

## Identifying promoters

```{r}
## Assign the promoter annotation to a variable
promoterAnnotation <- promoterAnnotation_gencode_v19

## Run the proActiv function to analyze promoter activity
result <- proActiv(
    files = files, 
    promoterAnnotation = promoterAnnotation, 
    condition = condition
)

## Display the result object
show(result)
```

```{r}
# Remove single-exon transcripts/promoters by eliminating entries where promoter counts are NA
final_result <- result[complete.cases(assays(result)$promoterCounts), ]

# Convert rowData to a data frame
row_data <- rowData(final_result)
promoter_df <- as.data.frame(row_data)

# Display the first few rows of the data frame for a preview
head(promoter_df)

# Ensure 'seqnames' is a factor with a specific order: chromosomes 1 through 22, followed by 'chrX', 'chrY', and 'chrM'
promoter_df$seqnames <- factor(promoter_df$seqnames, 
                               levels = c(paste0("chr", 1:22), "chrX", "chrY", "chrM"))

# Sort the data frame first by 'seqnames' and then by 'start' within each 'seqnames' group
promoter_df_sorted <- promoter_df[order(promoter_df$seqnames, promoter_df$start), ]

# Convert list columns in the data frame to character vectors
# Lists are flattened into comma-separated strings for each row
promoter_df_sorted[] <- lapply(promoter_df_sorted, function(x) {
  if (is.list(x)) {
    sapply(x, function(y) paste(unlist(y), collapse = ", "))
  } else {
    x
  }
})

# Filter rows where 'internalPromoter' is FALSE or NA, or 'seqnames' is NA
external_promoter_df_sorted <- promoter_df_sorted[
  promoter_df_sorted$internalPromoter == FALSE | is.na(promoter_df_sorted$internalPromoter) | is.na(promoter_df_sorted$seqnames), ]

# Save the filtered data frame to a CSV file
write.table(external_promoter_df_sorted, 
            file = "/mnt/NAS4/user_data/np-nhu/rna_human/external_promoter_17092024_c25.csv", 
            sep = ",", row.names = FALSE, quote = FALSE)

# Extract 'seqnames' and 'start' columns for the BED file
external_promoter_coordinated <- external_promoter_df_sorted[ , c("seqnames", "start")]

# Save the filtered promoter coordinates to a BED file without row names and headers
write.table(external_promoter_coordinated, 
            file = "/mnt/NAS4/user_data/np-nhu/rna_human/extermal_promoter_region_17092024_c25.bed", 
            quote = FALSE, row.names = FALSE, col.names = FALSE, sep = "\t")
```

### Promoter category proportion

```{r}
rdata <- rowData(result)
## Create a long dataframe summarizing cell line and promoter class
pdata1 <- data.frame(cellLine = rep(c('HIGH', 'LOW'), each = nrow(rdata)),
                       promoterClass = as.factor(c(rdata$HIGH.class, rdata$LOW.class)))

ggplot(na.omit(pdata1)) +
  geom_bar(aes(x = cellLine, fill = promoterClass)) + 
  xlab('Cell Lines') + ylab('Count') +  labs(fill = 'Promoter Category') +
  ggtitle('Categorization of Promoters')
```

### t-SNE plot with all active promoters

```{r}
## Remove inactive promoters (sparse rows)
data <- assays(result)$absolutePromoterActivity %>% dplyr::filter(rowSums(.) > 0)
data <- data.frame(t(data))
data
data$Sample <- as.factor(condition)

set.seed(40) # for reproducibility

# Perform t-SNE
tsne.out <- Rtsne(as.matrix(subset(data, select = -c(Sample))), perplexity = 1)
tsne.out

# Plot the t-SNE result
plot(x = tsne.out$Y[,1], y = tsne.out$Y[,2], 
     bg = data$Sample, asp = 1,
     col = 'black', pch = 24, cex = 4,
     main = 't-SNE plot with promoters \n active in at least one sample',
     xlab = 'T-SNE1', ylab = 'T-SNE2',
     xlim = c(-150,150), ylim = c(-150,150))

legend('topright', inset = .02, title = 'Cell Lines',
       legend = unique(condition), pch = 24, 
       pt.bg = 1:length(unique(condition)), cex = 1.5, bty = 'n')
```

## Identifying alternative promoters

```{r}
alternativePromoters <- getAlternativePromoters(result = result, referenceCondition = "HIGH")
show(alternativePromoters)

```

```{r}
# Combine upReg and downReg into a single data frame
alternativePromoters_combined <- rbind(alternativePromoters$upReg, alternativePromoters$downReg)

# Assume row_data_df contains columns promoterId and geneId
combine_alternative_promoter_df <- promoter_df %>%
  dplyr::filter(promoterId %in% alternativePromoters_combined$promoterId & 
                geneId %in% alternativePromoters_combined$geneId)

# Display the result
combine_alternative_promoter_df[, c(1, 2, 3, 9:14)]
combine_alternative_promoter_df
```

### Gene information

```{r}
# Load necessary libraries at once
library(biomaRt)
library(dplyr)

# Assuming combine_alternative_promoter_df is your data frame containing geneId
# Clean geneId by removing version suffix (e.g., '.11', '.12', etc.)
combine_alternative_promoter_df <- combine_alternative_promoter_df %>%
  mutate(geneId = gsub("\\..*", "", geneId))

# Connect to the Ensembl GRCh37 database using biomaRt
ensembl37 <- useMart(
  host = "https://grch37.ensembl.org", 
  biomart = "ENSEMBL_MART_ENSEMBL", 
  dataset = "hsapiens_gene_ensembl"
)

# Query the Ensembl database for relevant gene information
gene_info <- getBM(
  attributes = c("ensembl_gene_id", "external_gene_name", "chromosome_name", 
                 "start_position", "end_position", "gene_biotype", "description"),
  filters = "ensembl_gene_id", 
  values = combine_alternative_promoter_df$geneId, 
  mart = ensembl37
)

# Factor chromosome names with levels in the correct order and sort the data
gene_info <- gene_info %>%
  mutate(chromosome_name = factor(chromosome_name, levels = c(as.character(1:22), "X", "Y", "MT"))) %>%
  arrange(chromosome_name, start_position)

# Merge the two dataframes on the cleaned geneId
merged_df <- combine_alternative_promoter_df %>%
  left_join(gene_info, by = c("geneId" = "ensembl_gene_id"))

# Convert list columns in the data frame to character vectors
# Lists are flattened into comma-separated strings for each row
merged_df[] <- lapply(merged_df, function(x) {
  if (is.list(x)) {
    sapply(x, function(y) paste(unlist(y), collapse = ", "))
  } else {
    x
  }
})

# Save the sorted data frame to a CSV file
write.table(merged_df, 
            file = "/mnt/NAS4/user_data/np-nhu/rna_human//alternative_promoter_17092024_c25.csv", 
            sep = ",", row.names = FALSE, quote = FALSE)
```

### The presence of alternative promoters increases in high-grade samples.

```{r}
plots_up <- boxplotPromoters(result, "ENSG00000130304")

# Boxplot of absolute promoter activity
grid.arrange(plots_up[[1]], plots_up[[3]], nrow = 1, ncol = 2, widths = c(3, 2))
```

### The presence of alternative promoters decreases in high-grade samples.

```{r}
plots_down <- boxplotPromoters(result, "ENSG00000159692.11")

# Boxplot of absolute promoter activity
grid.arrange(plots_down[[1]], plots_down[[3]], nrow = 1, ncol = 2, widths = c(3, 2))
```
